<!--------------------------------------------------------------------------------------------->
<!--                                                                                         -->
<!--                                     Data exploration                                    -->
<!--                                                                                         -->
<!--------------------------------------------------------------------------------------------->

<!-- Loading useful libraries --->

```{r, include = F}
library(sva)
```

# **Data exploration**

<!----------------------------------------------------->
<!--                                                 -->
<!--         Principal component analysis            -->
<!--                                                 -->
<!----------------------------------------------------->

## **Principal component analysis (PCA)**
  
:::info  
Principal component analysis (PCA) allows a representation of the samples in a low dimensional space estimated considering all genes' expressions (after background genes filtering)
:::

<!-- Calculate PCA   -->
  
```{r pca}
pca_results <- compute_pca(eset_count_filtered_vst)
```


### PCA evaluations {.tabset}

#### Scree-plot

<!-- Scree-plot (eigenvalues) -->

```{r pca-details, fig.width=5, fig.height=5}
draw_pca_eigenvalues(pca_results)
```

#### Loadings

<!-- loadings  -->
<!-- For large datasets, one has to select the amount of features to plot with select.var -->

```{r, fig.width=5, fig.height=5}
draw_pca_loadings(pca_results, select.var = list(contrib = 2))
```


### 2D-plots

<!-- 2D plots displayed side by side  -->

```{r pca-graph1, fig.width=12, fig.height=5}

g1 <- draw_pca_individuals(
  eset_count_filtered_vst,
  pca_results,
  col_color = "Donor",
  palette = "nejm",
  add_ellipse = TRUE,
  title = "PCA of the samples by Donor"
  ) 

g2 <- draw_pca_individuals(
  eset_count_filtered_vst,
  pca_results,
  col_color = "Treatment",
  palette = "dark2",
  add_ellipse = TRUE,
  title = "PCA of the samples by Treatment",
  shape_manual = c(5, 6, 7, 8)
  ) 


g3 <- draw_pca_individuals(
  eset_count_filtered_vst,
  pca_results,
  col_color = "RIN",
  palette = "spectral",
  add_ellipse = FALSE,
  title = "PCA of the samples by RIN"
)

ggpubr::ggarrange(g1, g2, g3,
                  ncol = 3,
                  common.legend = FALSE,
                  legend = "bottom")
```


### 2D-interactive plots {.tabset}

<!-- 2D interactive plots displayed in different tabs  -->

#### PCA by Treatment

```{r pca-graph2, fig.width = 7, fig.height = 5}
draw_pca_individuals(
  eset_count_filtered_vst,
  pca_results,
  col_color = "Treatment",
  palette = "dark2",
  add_ellipse = TRUE,
  title = "PCA of the samples by Treatment",
  interactive = TRUE, 
  col_hover_info= c("Treatment", "Donor")
)  
```

#### PCA by RIN

```{r pca-graph3, fig.width = 7, fig.height = 5}
draw_pca_individuals(
  eset_count_filtered_vst,
  pca_results,
  col_color = "RIN",
  palette = "spectral",
  add_ellipse = FALSE,
  title = "PCA of the samples by RIN",
  interactive = TRUE
  )  
```
  
  
### 3D-interactive plots {.tabset}

<!-- 3D interactive plots displayed in different tabs  -->

#### PCA by Treatment

```{r pca-graph4, fig.width = 8, fig.height = 5}

draw_pca_individuals(
  eset_count_filtered_vst,
  pca_results,
  col_color = "Treatment",
  palette = "dark2",
  title = "PCA of the samples by Treatment",
  interactive = TRUE, 
  comp = 1:3,
  add_ellipse = TRUE,
  col_hover_info= c("Treatment", "Donor")
) 
```

#### PCA by RIN

```{r pca-graph5, fig.width = 8, fig.height = 5}

draw_pca_individuals(
  eset_count_filtered_vst,
  pca_results,
  col_color = "RIN",
  palette = "spectral",
  title = "PCA of the samples by RIN",
  comp = 1:3,
  interactive = TRUE
  ) 
```

###

:::conclusion
+ Treatment effect is strong only for some compounds

+ There is a donor effect that will be accounted for in the differential expression analysis
:::



<!----------------------------------------------------->
<!--                                                 -->
<!--  Partial Least Squares - Discriminant Analysis  -->
<!--                                                 -->
<!----------------------------------------------------->

## **Partial least squares - discriminant analysis (PLS-DA)**
  
:::info  
Partial least squares - discriminant analysis (PLS-DA) allows a representation of the samples in a low dimensional space estimated considering genes that are selected as the most discriminant between groups of a variable of interest 
:::
  
<!-- Calculate PLS-DA   -->

```{r plsda, include = FALSE}
plsda_results <- compute_plsda(eset_count_filtered_vst, y_outcome_column = "Treatment")

plsda_results_multi <- compute_plsda(eset_count_filtered_vst, y_outcome_column = "Treatment", 
                                     multilevel = "Donor")
```

### Evaluation of the PLS-DA {.tabset}

<!-- Loadings and variable plots  -->

#### Variables

```{r plsda-details, fig.width=8, fig.height=5}
draw_plsda_loadings(
  plsda_results,
  eset_count_filtered_vst,
  title = "PLSDA variable representation",
  comp = 1:2,
  gene_label_column = "GeneID",
  rad.in = 0.9,
  cutoff = 0.8,
  ndisplay = 50,
  interactive = FALSE
) + coord_equal() 
```

#### Importance

```{r plsda-details-2, fig.width=8, fig.height=5}
draw_plsda_importance(
  plsda_results,
  eset_count_filtered_vst,
  title = "PLSDA scatter plot for individuals",
  comp = 1,
  ndisplay = 20,
  gene_label_column = "GeneID",
  contrib = 'max',
  method = 'mean'
)
```

#### Performances

```{r plsda-details-3, fig.width=8, fig.height=5, eval = FALSE}
plot(mixOmics::perf(
    plsda_results,
    validation = "Mfold",
    folds = 5,
    nrepeat = 10,
    progressBar = FALSE
  ))
```


### 2D-plot 

<!-- 2D plots displayed side by side  -->

```{r plsda-graph,  fig.width=9, fig.height=5}

g1 <- draw_plsda_individuals(
  eset_count_filtered_vst,
  plsda_results,
  add_ellipse = TRUE,
  comp = 1:2,
  palette = "dark2",
  title = "PLSDA of the samples by Treatment"
)  

g2 <- draw_plsda_individuals(
  eset_count_filtered_vst,
  plsda_results_multi,
  add_ellipse = TRUE,
  comp = 1:2,
  palette = "dark2",
  title = "PLSDA of the samples by Treatment \nleveled by Donor"
)  

ggpubr::ggarrange(g1,
                  g2,
                  common.legend = T,
                  legend = "bottom")

```


### 2D-interactive plots {.tabset}

<!-- 2D interactive plots displayed in different tabs  -->

#### PLS-DA by Treament

```{r plsda-graph-2, fig.width = 8, fig.height = 5}

draw_plsda_individuals(
  eset_count_filtered_vst,
  plsda_results,
  add_ellipse = TRUE,
  comp = 1:2,
  palette = "dark2",
  title = "PLSDA of the samples by Treatment", 
  interactive = TRUE
)  

```
  
#### PLS-DA by Treament levelled by Donor

```{r plsda-graph-3, fig.width = 8, fig.height = 5}
draw_plsda_individuals(
  eset_count_filtered_vst,
  plsda_results_multi,
  add_ellipse = TRUE,
  comp = 1:2,
  palette = "dark2",
  title = "PLSDA of the samples by Treatment \nleveled by Donor",
  interactive = TRUE
)  

```


### 3D-interactive plots {.tabset}

<!-- 2D interactive plots displayed in different tabs  -->

#### PLS-DA by Treament
```{r plsda-graph-4, fig.width = 8, fig.height = 5}

draw_plsda_individuals(
  eset_count_filtered_vst,
  plsda_results,
  comp = 1:3,
  palette = "dark2",
  add_ellipse = TRUE,
  title = "PLSDA of the samples by Treatment", 
  interactive = TRUE
)

```
  
#### PLS-DA by Treament levelled by Donor

```{r plsda-graph-5, fig.width = 8, fig.height = 5}
draw_plsda_individuals(
  eset_count_filtered_vst,
  plsda_results_multi,
  comp = 1:3,
  palette = "dark2",
  title = "PLSDA of the samples by Treatment \nleveled by Donor",
  subtitle = "leveled by Donor", 
  interactive = TRUE
)

```
  
###

:::conclusion

+ Treatment effect is more noticeable after gene selection and accounting for donor effect

:::

<!----------------------------------------------------->
<!--                                                 -->
<!--                 Sample Clustering               -->
<!--                                                 -->
<!----------------------------------------------------->

## **Sample clustering**
  
<!-- Heatmap of the similarity matrix calculated between samples  -->

```{r sample-clustering, fig.width=5, fig.height=5}
# Create a DESeq object dds to be used for clustering
dds <-
  DESeq2::DESeqDataSetFromMatrix(
    countData = exprs(eset_count_filtered),
    colData = pData(eset_count_filtered_vst),
    design = ~ Treatment + Donor
  )

# Caculation of vst normalized expression according to a provided model
vsd <- DESeq2::vst(dds, blind = FALSE)

# Calculation of an Euclidean distance matrix between samples
sampleDists <- SummarizedExperiment::assay(vsd) %>% t() %>% dist()

# Formatting distance matrix for clustering
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(vsd$Treatment, vsd$Donor, sep = "_")
colnames(sampleDistMatrix) <- paste(vsd$Treatment, vsd$Donor, sep = "_")
colors <-
  colorRampPalette(rev(RColorBrewer::brewer.pal(9, "Blues")))(255)

# Sample by sample heatmap and clustering
pheatmap::pheatmap(
  sampleDistMatrix,
  clustering_distance_rows = sampleDists,
  show_rownames = T,
  clustering_distance_cols = sampleDists,
  col = colors)

```


:::warning
+ Strong donor effect confirmed by clustering
:::

<!----------------------------------------------------->
<!--                                                 -->
<!--              Batch effect correction            -->
<!--                                                 -->
<!----------------------------------------------------->

## **Batch effect correction**

The Donor effect is strong and can be removed from the data before downstream exploratory analyses


<!---------------------------------------------------------------------->
<!--     1st approach using ComBat for a single categorical batch     -->
<!---------------------------------------------------------------------->
<!------------------->
<!-- Known batches -->
<!------------------->

### **Combat**
                        
```{r ComBat, include = FALSE}

# Reference model for ComBat including the variables to be preserved (e.g variable of interests)
mod_combat <-
  model.matrix( ~ 1 + Treatment, data = pData(eset_count_filtered_vst))

# Applying ComBat
count_filtered_vst_combat <-
  sva::ComBat(
    dat = eset_count_filtered_vst %>% exprs %>% as.matrix,
    batch = eset_count_filtered_vst %>% pData %>% dplyr::pull(Donor) %>% as.character,
    mod = mod_combat,
    par.prior = TRUE,
    prior.plots = FALSE
  )

# Create normalized eSet
eset_count_filtered_vst_combat <-
  eset_object(
    expression_data = count_filtered_vst_combat,
    phenotype_data = pData(eset_count_filtered),
    feature_data = fData(eset_count_filtered)
  )

```

<!-- Comparison of PCA before and after batch removal -->

```{r ComBat-PCA,  fig.width=11, fig.height=5}

# Calulates PCA before and after batch removal
pca_results <- compute_pca(eset_count_filtered_vst)
pca_results_combat <- compute_pca(eset_count_filtered_vst_combat)

# PCA plots
d1 <- draw_pca_individuals(
  eset_count_filtered_vst,
  pca_results,
  col_color = "Treatment",
  palette = "dark2",
  add_ellipse = TRUE,
  title = "PCA of the samples before batch correction",
)  

d2 <- draw_pca_individuals(
  eset_count_filtered_vst_combat,
  pca_results_combat,
  col_color = "Treatment",
  palette = "dark2",
  add_ellipse = TRUE,
  title = "PCA of the samples after batch correction",
)  

ggpubr::ggarrange(d1, d2,
                  ncol = 2 ,
                  common.legend = T,
                  legend = "bottom")
```


:::conclusion
+ Batch effect correction is quite well taken into account 
:::


<!---------------------------------------------------------------------->
<!-- 2nd approach using Residuals for any number and types of batches -->
<!---------------------------------------------------------------------->
<!------------------->
<!-- Known batches -->
<!------------------->

### **GLM Residuals**


```{r residuals}

# This model is passed to the linear model and corresponds to the variables one wants to remove the effects
mod_residual <- model.matrix( ~ 1 + Donor, data = pData(eset_count_filtered))

# Vst normalization 
dds <- DESeq2::DESeqDataSetFromMatrix(
  countData = exprs(eset_count_filtered),
  colData = pData(eset_count_filtered),
  design = ~ 1 + Treatment # This model is the model used for the normalization vst and can include the variable of interest
) 

dds.vst  <- SummarizedExperiment::assay(dds %>% DESeq2::vst(blind = F))

# Run of Limma linear regression model
dds.limma.fit <- limma::lmFit(dds.vst, design = mod_residual)

# Residuals of previous model corresponds to the normalized matrix
resid <- residuals(dds.limma.fit, y = dds.vst)

# Creation a a normalized eSet
eset_count_filtered_vst_residuals <-
  eset_object(
    expression_data = resid,
    phenotype_data = pData(eset_count_filtered),
    feature_data = fData(eset_count_filtered)
  )

```

<!-- Comparison of PCA before and after batch removal -->

```{r residuals-PCA,  fig.width=11, fig.height=5}

pca_results <- compute_pca(eset_count_filtered_vst)
pca_results_residuals <- compute_pca(eset_count_filtered_vst_residuals)

d1 <- draw_pca_individuals(
  eset_count_filtered_vst,
  pca_results,
  col_color = "Treatment",
  palette = "dark2",
  add_ellipse = TRUE,
  title = "PCA of the samples before batch correction",
)  

d2 <- draw_pca_individuals(
  eset_count_filtered_vst_residuals,
  pca_results_residuals,
  col_color = "Treatment",
  palette = "dark2",
  add_ellipse = TRUE,
  title = "PCA of the samples after batch correction",
)  

ggpubr::ggarrange(d1, d2,
                  ncol = 2 ,
                  common.legend = TRUE,
                  legend = "bottom")
```

:::conclusion
+ Batch effect correction is quite well taken into account 
:::

<!---------------------------------------------------------------------->
<!--   3rd approach using SVA/RUV for a single categorical batch      -->
<!---------------------------------------------------------------------->
<!--------------------->
<!-- Unknown batches -->
<!--------------------->
                        
### **Surrogate Variable Analysis (SVA)**

<!-- Surrogate variable analysis (SVA) -->

```{r sva, message = FALSE, include = FALSE}
# Full model including all variables of interest (that we know have an effect on the outcome that we want to estimate)
mod_sva <-
  model.matrix( ~ 1 + Treatment, data = pData(eset_count_filtered_vst))

# Null model with only an intercept
mod_null <-  model.matrix( ~ 1, data = pData(eset_count_filtered_vst))

# Estimation of the number of surogate variables
n_sv = num.sv(eset_count_filtered_vst %>% exprs , mod_sva, method = "be")

# Calculation of the surogate variables
svobj = sva(eset_count_filtered_vst %>% exprs, mod_sva, mod_null, n.sv = n_sv)

# svobj$sv contains the surrogate variables that can be used as adjustments in DEA models or as known batch variables in the 2nd approach using Residuals to obtain a batch free eSet
```

#### **Surrogate variables evaluation through PCA** 

```{r sva-eval, fig.width=11, fig.height=5}
# Creation of a bogus pca_object to plot the surrogate variables
pcs <- svobj$sv
colnames(pcs) <- paste0("Dim.", 1:ncol(pcs))
rownames(pcs) <- colnames(eset_count_filtered_vst)

eig <- data.frame(eigenvalue = rep(0, n_sv))
eig <- eig %>% mutate(var = 100*eigenvalue/sum(eigenvalue)) %>% mutate(cumvar = cumsum(var))
colnames(eig) <- c("eigenvalue", "percentage of variance", "cumulative percentage of variance")
rownames(eig) <- paste0("comp ", 1:nrow(eig))

# Creation of a bogus PCA object
pca_object <- list(ind = list(coord = pcs), eig = eig)

# PCA plot of the SV
d1 <- draw_pca_individuals(
  eset_count_filtered_vst,
  pca_object,
  col_color = "Treatment",
  palette = "dark2",
  add_ellipse = TRUE,
  title = "Sample plot on the Surrogate Variable spaces",
)

d2 <- draw_pca_individuals(
  eset_count_filtered_vst,
  pca_object,
  col_color = "Donor",
  palette = "dark2",
  add_ellipse = TRUE,
  title = "Sample plot on the Surrogate Variable spaces",
)

ggpubr::ggarrange(d1,
                  d2,
                  ncol = 2 ,
                  common.legend = F,
                  legend = "bottom")
```

:::conclusion
+ The first two surrogate variables corresponds to the donor groups and can be used in downstream analyses 
:::

<!-- Remove unwanted variation (RUV) -->
