<!--------------------------------------------------------------------------------------------->
<!--                                                                                         -->
<!--                            Differential Expression analysis                             -->
<!--                                                                                         -->
<!--------------------------------------------------------------------------------------------->


# **Differential expression analysis**

## **Statistical modeling**

:::info
+ We consider the statistical statistical model

$$Expression \sim \textit{Treatment + Donor}  $$

+ We use the R package `DESeq` to perform the differential expression analysis

+ Counts are directly modeled through a Negative Binomial distribution and fitted using a Generalized Linear Model (GLM). An independent filtering is then used to estimate significant DEGs.

+ Fold-change (FC) and adjusted p-values (FDR of Benjamini-Hochberg correction) are used to identify differentially expressed genes (DEG)

+ We consider significant genes with a $|FC| > 1.3$ and a $p_{adj} < 0.05$ (FDR)

:::



<!-- This script performs DESeq analysis                                                     -->
<!-- It produces a tibble with, for each contrast, its table of results and table of DEGs    -->
<!-- This tibble is then used in the following script for result exploration                 -->        


<!-- Loading useful libraries --->

```{r, echo=FALSE}
library(purrr)
```

<!-- Setting FC and FDR cut-offs  -->
```{r Global-variables-1}
t_pvalue <- 0.05
t_FC <- 1.3
```

<!-- /!\ The variables of interest, used to define the contrasts, have to be factors   -->

```{r DESeq2-no-cov}
# Model without covariate
model <- formula(~  Treatment)

# Model with covariates (Treatment preferably last)
model <- formula(~ Donor + Treatment )

# Contrasts
contr_list = list(c("Treatment", "AAA", "Placebo"),
                     c("Treatment", "BBB", "Placebo"),
                     c("Treatment", "CCC", "Placebo"))

# Perform differential expression analysis with DESeq
DEA_res <- dea_deseq(eset_object = eset_count_filtered,
           model = model,
           contr_list = contr_list, 
           feature_colname = "Genes")

# Add the tables of DEGs to the results
DEA_res <- DEA_res %>% mutate(DEgenes = purrr::map(data,
                       ~ .x %>% subset_deg(., "FC", t_FC, "FDR", t_pvalue, order = "FC")))

```

