<!--------------------------------------------------------------------------------------------->
<!--                                                                                         -->
<!--                                    Results Exploration                                  -->
<!--                                                                                         -->
<!--------------------------------------------------------------------------------------------->

<!-- This script proposes several tables an graphics to describe the results                 -->
<!-- It handles all contrasts simultaneously                                                 -->
<!-- Results can be resented with a tab per contrast or with join plots                      -->   
<!-- User can decide of its representation type and is encouraged to propose new ones        -->   


<!-- Loading useful libraries --->
```{r}
library(purrr)
library(knitr)
library(flextable)
library(ggpubr)
library(tidyr)
```

## **Results exploration**

### **Summary of DEGs (on `r nrow(DEA_res$data[[1]])` genes)** {.tabset}

<!-- Creates a tab with the summary for each contrast  -->

```{r, results= 'asis', echo = FALSE, eval= TRUE}
cat("\n")
purrr::pwalk(DEA_res,
             function(data, contrast,  ...) {
               
               cat(paste0("  \n#### ", contrast, " \n"))
               cat("\n")
               cat(knitr::knit_print(
                 table_deg_significant(
                   result_table = data,
                   foldchange_column = "FC",
                   foldchange_threshold = c(1.3, 1.5, 2),
                   padj_column = "FDR",
                   padj_threshold = c(0.05, 0.1, 0.2),
                   interactive = interactive
                 )))
              })
```

:::conclusion
+ AAA and CCC produce nice gene signatures

+ BBB has no DEGs whatever the thresholds
:::

### **p-value distribution**

<!-- Creates a plot with all p-value histograms -->

```{r, fig.width=15, fig.height=5, fig.fullwidth=TRUE}
pmap(DEA_res, function(data, contrast, ...) {
  draw_pvalue_histogram(
    data,
    "pvalue",
    pi0_estimate = TRUE,
    alpha = 0.05
  )  
  
})  %>% ggpubr::ggarrange(
  plotlist = .,
  nrow = 1,
  labels = paste(DEA_res$contrast),
  common.legend = TRUE,
  legend = "bottom",
  vjust = 1.2
) %>%   print()

```

:::conclusion
+ p-value distributions of AAA and CCC show important signals but not for BBB
:::



### **Table of results** {.tabset}

<!-- Creates a tab with the table for each contrast  -->

```{r, results='asis', fig.width= 12, fig.height=8 , eval=TRUE, echo=FALSE}
purrr::pwalk(DEA_res,
             function(data, contrast, DEgenes, ...) {
               cat("  \n  \n")
               cat(paste0(
                 "  \n#### ",
                 contrast,
                 " (",
                 nrow(DEgenes),
                 " genes differentially expressed) \n"
               ))

               # if (nrow(DEgenes) > 0) {
                 cat("\n")
                 cat("::::mycolumns")
                 cat("\n")
                 cat(":::mycolumn")
                 cat("\n")
                 cat(knitr::knit_print(
                   display_table(
                     DEgenes,
                     interactive = interactive, 
                     # color_shade_column = "FDR",
                     bar_column = c("FDR","FC"),
                     highlighted_row = FC > 10 | FC < -10,
                     highlighted_color = "lightblue",
                     nmax_display = 10
                   )
                 ))

                 cat("\n")
                 cat(":::")
                 cat("\n")
                 cat("::::")
                 cat("\n")
               # }
               cat("\n")

             })

```

### **Table of results for all contrasts** 

<!-- Creates a table of results for all contrasts  -->

::::mycolumns

:::mycolumn

```{r}

# Create a result table with different contrasts dysplayed in columns
fulltable <-
  DEA_res %>% dplyr::select(contrast, data) %>% tidyr::unnest() %>%
  tidyr::pivot_wider(
    id_cols = "Genes",
    names_from = "contrast",
    values_from = c("FC", "FDR")
  ) %>% mutate(across(starts_with("FC_"), ~ sign(.) * log2(abs(.)), .names = "log2{.col}")) # to get log2 FC as well

display_table(
  fulltable,
  interactive = TRUE,
  # color_shade_column = "FDR",
  bar_column = paste0("FC_", DEA_res$contrast),
  highlighted_color = "lightblue",
  nmax_display = 10
)

```

:::

::::

###


### **Volcano plots** {.tabset}

<!-- Creates a tab with the plot for each contrast  -->

```{r, results='asis', fig.width= 8, fig.height=6}
purrr::pwalk(DEA_res,
             function(data, contrast, ...) {
               cat(paste0("  \n#### ", contrast , "  \n"))
               draw_volcano(
                 data = data,
                 col_log2fc = "log2FC",
                 col_signif = "pvalue",
                 col_label = "Genes",
                 fc_cutoff = 1.3,
                 signif_cutoff = 0.05,
                 type = c("pv"),
                 labelled = data %>% pull(Genes) %>% head(100),
                 #focused = c("C1QA", "C1QB", "C1QC"),
                 shape = c(17, 1, 16),
                 alpha = c(0.9, 0.4, 0.7),
                 size = c(5, 2, 2),
                 legend = TRUE
                 )   %>% print()
               cat("\n")
               
             })

  
```

###

:::info
+ CCC has mostly down-regulated genes compared to AAA that has both

+ FC are higher for CCC 
:::



### **Heatmap**  {.tabset}

```{r, fig.width= 8, fig.height=7}
# Select gene list for heatmap
gene_list = DEA_res$DEgenes %>% purrr::map( ~ .x) %>% bind_rows()  %>%  pull(Genes)

# Plot
draw_heatmap(
  filter_features_from_values(
    eset_count_filtered_log2,
    values = gene_list,
    column = "GeneID"
  ),
  column_annotation_names = c("Treatment", "Donor"),
  column_palette_selection = c("dark2", "nejm"),
  show_row_names = FALSE,
  name = "Exprs",
  scale_data = TRUE,
  column_title = "Heatmap of expression for genes DE in at least one contrast", 
  # cluster_rows = FALSE,
  # cluster_columns = FALSE,
  interactive = FALSE
)
```

###

:::conclusion
+ BBB and placebo have opposite effect compared to AAA and CCC on DEGs
:::


### **Gene visualization**

#### Box plots top 10 Up-regulated DEG {.tabset}

<!-- Creates a tab with the plots for each contrast  -->

```{r, results='asis', fig.width=12, fig.height=8 , warnings = FALSE}
purrr::pwalk(DEA_res,
             function(DEgenes, contrasts, ...) {
               cat("  \n  \n")
               cat(paste0("  \n##### ",
                          contrasts,
                          "\n"))
               if (nrow(DEgenes) > 0) {
                 boxplottop20 <-
                   purrr::map(
                     DEgenes %>% filter(FC >0) %>% slice_max(FC, n = 10) %>% pull(Genes),
                     ~
                       draw_boxplot((
                         eset_tpm %>% filter_features_from_values(values = .x, column = "GeneID")
                       ),
                       col_x = "Treatment",
                       col_y = .x,
                       col_color = "Treatment",
                       y_transformation  = ~log2(. + 1),
                       palette  = "dark2",
                       hline_threshold  = log2(esti_tpm + 1), 
                       title = .x,
                       add_point = "jitter_on_x"
                       )) %>% 
                   ggpubr::ggarrange(plotlist = .,
                                     nrow = ifelse(nrow(DEgenes) < 5, 1, 2),
                                     ncol = min(5, nrow(DEgenes)),
                                     common.legend = TRUE,
                                     legend = "bottom") %>%
                   print()
                 cat("\n")

               }
               cat("\n")
             })
```



#### Box plots top 10 Down-regulated DEG {.tabset}

<!-- Creates a tab with the plots for each contrast  -->

```{r, results='asis', fig.width=12, fig.height=8 , warnings = FALSE}
purrr::pwalk(DEA_res,
             function(DEgenes, contrasts, ...) {
               cat("  \n  \n")
               cat(paste0("  \n##### ",
                          contrasts,
                          "\n"))
               if (nrow(DEgenes) > 0) {
                 boxplottop20 <-
                purrr::map(
                     DEgenes %>% 
                       filter(FC <0) %>% 
                       slice_min(FC, n = 10) %>% 
                       pull(Genes),
                     ~ draw_boxplot((eset_tpm %>% 
                                       filter_features_from_values(values = .x, 
                                                                   column = "GeneID")),
                                    col_x = "Treatment",
                                    col_y = .x,
                                    col_color = "Treatment",
                                    y_transformation  = ~log2(. + 1),
                                    palette  = "dark2",
                                    hline_threshold  = log2(esti_tpm + 1), 
                                    title = .x,
                                    add_point = "jitter_on_x")) %>% 
                   ggpubr::ggarrange(
                     plotlist = .,
                     nrow = ifelse(nrow(DEgenes) < 5, 1, 2),
                     ncol = min(5, nrow(DEgenes)),
                     common.legend = TRUE,
                     legend = "bottom"
                   ) %>%   
                   print()
                 cat("\n")
               }
               cat("\n")
             })
```


#### Focus on a particular gene {.tabset}

##### Static version with pvalues

```{r, fig.width=7, fig.height=5}
# Select gene of interest
gene_interest <- "CSTA"

# Get stat and pv from DEA
stat <- 
  pmap(DEA_res, 
       function(data, contrast, ...) {
         data %>% 
           filter(., Genes == gene_interest) %>% 
           select(., c(Genes, FC, FDR)) %>% mutate(contrast = contrast)}) %>% 
  bind_rows() %>% 
  separate(contrast, c("group1", "group2"), sep = "_vs_")


# Plot
draw_boxplot(
  eset_tpm %>%
    filter_features_from_values(values = gene_interest,
                                column = "GeneID"),
  col_x = "Treatment",
  col_y = gene_interest,
  col_color = "Treatment",
  y_transformation  = ~ log2(. + 1),
  palette  = "dark2",
  hline_threshold  = log2(esti_tpm + 1),
  title = "CSTA gene",
  add_point = "jitter_on_x",
  stat_to_display = stat) + 
  labs(title = "CSTA gene")

```

##### Interactive version without pvalues

```{r, fig.width=7, fig.height=5}
# draw_boxplot(
#   eset_tpm %>% 
#     genes_filter_fromlist(gene_list = gene_interest, 
#                           colname_fdata = "GeneID"),
# x = "Treatment",
# y_filter = gene_interest,
# exprs_trans = ~log2(. + 1),
#   y_unit = "tpm",
# palette = "Dark2",
# background_threshold = log2(esti_tpm + 1),
# background_threshold_name = "           Cutoff",
# title = "C1QA gene",
# interactive = TRUE
# ) 

```



#### **Profile plots** {.tabset}

##### Static version

```{r, fig.width=7, fig.height=5}
draw_profile(
  eset_tpm,
  col_y = gene_interest,
  col_x = "Treatment",
  col_group = "Donor",
  col_color = "Donor",
  palette  = "accent",
  point_size = 1,
  hline_threshold  = log2(esti_tpm + 1),
  y_transformation  = ~ log2(. + 1)) + 
  labs(title = "CSTA gene")

```

##### Interactive version

```{r, fig.width=7, fig.height=5}
# draw_profileplot(eset_tpm,
#                  x = "Treatment",
#                  y_filter ="A1BG",
#                  groupby = "Donor",
#                  y_unit = "tpm",
#                  exprs_trans = ~log2(. + 1),
#                  palette = "Dark2",
#                  background_threshold = NULL,
#                  interactive = TRUE)
```

