<!--------------------------------------------------------------------------------------------->
<!--                                                                                         -->
<!--                                   Enrichment analysis                                   -->
<!--                                                                                         -->
<!--------------------------------------------------------------------------------------------->

```{r, echo = FALSE}
library(dplyr)
library(msigdbr)
library(clusterProfiler)
library(fgsea)
library(enrichplot)
library(ggplot2)
library(ggstance)
library(forcats)

# load("/HPC_projets/pexmvd/mubz_cb/Methodo/BMK_packages/DEAres.RData")

```

# **Enrichment Analysis**

<!--------------------------------------------------------------->
<!--                                                           -->
<!--                  Pathway database selection               -->
<!--                                                           -->
<!--------------------------------------------------------------->

```{r}
# mSigDB
m_df = msigdbr(species = "Homo sapiens", category = "H")
m_list = m_df %>% split(x = .$gene_symbol, f = .$gs_name)

enrichDB <- m_df %>%
  dplyr::select(gs_name, gene_symbol)

# Reactome
# enrichDB <- msigdbr::msigdbr(species = "Homo sapiens",
#                  category = "C2",
#                  subcategory = "REACTOME") %>% select(gs_name, gene_symbol)

# KEGG
# enrichDB <- msigdbr::msigdbr(species = "Homo sapiens",
#                  category = "C2",
#                  subcategory = "KEGG") %>% select(gs_name, gene_symbol)

```

<!--------------------------------------------------------------->
<!--                                                           -->
<!--                       fGSEA apprpach                      -->
<!--                                                           -->
<!--------------------------------------------------------------->

## **fGSEA approach**


+ fGSEA enrichment analysis corresponds to a pre-ranked version of GSEA. A metric is selected and calculated for each gene that are then ranked accordingly. This metric is then passed to a classical GSEA analysis. An normalized enrichment score (NES) is provided for each pathway along with useful indicators helping the interpretation.

+ We consider a ranking metric for each gene that is based on the results for the differential expression

$$ sign(FC) \times(-log_{10}(\text{raw pvalue})) $$

+ The reference set of genes are the `r nrow(DEA_res$data[[1]])` genes that passed the pre-processing filters


+ Significant pathways are selected with a $FDR\leq 0.05$ and ordered by NES


```{r}
# Adding the fgseascores and results to the DEA_res object

# fGSEA scores
DEA_res <- DEA_res %>% mutate(fGSEA_scores = purrr::map(
  data,
  ~ .x %>% fgsea_ranked_score(
    deg_results = .,
    fc_label = "FC",
    pvalue_label = "pvalue",
    gene_label = "Genes"
  )
))


# Calculation of the completeness of each pathway for graphical representation
completnesse_data <- completeness_calculation(
  analyzed_data = fData(eset_count_filtered),
  gene_colname_data = "GeneID",
  reference_database = enrichDB,
  pathway_colname_reference = "gs_name",
  gene_colname_reference =  "gene_symbol"
)

# Running fGSEA

# IMPORTANT : fGSEA uses random walk and permutation procedures to estimate pvalues. It is necessary to set a seed to have reproducible results when runing several time this code

set.seed(1234)

# adding fgsea results object to DEA_res
DEA_res <- DEA_res %>% mutate(fGSEA_results = purrr::map(
  fGSEA_scores,
  ~ .x %>% fgsea_enrichment(
    statistics_score = .,
    pathways_database = enrichDB,
    pvalue_cutoff = 1
  ) 
))

# adding formatted fgsea results object to DEA_res

DEA_res <- 
  DEA_res %>% 
  mutate(fGSEA_results_4graph = purrr::map(
    fGSEA_results,
    ~ .x %>% res_format_enrichment(., type = "fgsea")
  )
)

DEA_res <- 
  DEA_res %>% 
  mutate(fGSEA_results_4graph = purrr::map(
    fGSEA_results_4graph,
    ~ .x %>% left_join(., completnesse_data)
  )
)

```

### **Table of results for all contrasts** 

<!-- Creates a table of results for all contrasts  -->

::::mycolumns

:::mycolumn

```{r}
# Full table of pathways for all contrasts

fulltable_pathways <-
  DEA_res %>% dplyr::select(contrast, fGSEA_results_4graph) %>% tidyr::unnest() %>%
  tidyr::pivot_wider(
    id_cols = c("ID", "pathway_size"),
    names_from = "contrast",
    values_from = c("NES", "FDR")
  )

# Display table
display_table(
  fulltable_pathways,
  interactive = TRUE,
  # color_shade_column = "FDR",
  bar_column = paste0("NES_", DEA_res$contrast),
  highlighted_color = "lightblue",
  nmax_display = 10
)
```

:::

::::


<!--------------------------------------------------------------->
<!--                                                           -->
<!--                  Pathways visualization                   -->
<!--                                                           -->
<!--------------------------------------------------------------->


### **Main significant pathways - barplots**  {.tabset}

:::info
+ The x-axis of the plot displays the NES

+ At the end of each bar is displayed the completeness that represents the amount of genes of the pathway present in the reference gene list

+ Completeness = #{genes in the pathway} / #{genes in the reference geneset}

:::

<!-- Creates a tab with the plot for each contrast  -->

```{r, results='asis', fig.width= 8, fig.height=6}
purrr::pwalk(DEA_res,
             function(fGSEA_results_4graph, contrast, ...) {
               cat(paste0("  \n#### ", contrast , "  \n"))
               draw_barplot_score(
                 enrichment_results = fGSEA_results_4graph,
                 xlabel = "NES",
                 pathway_colname = "ID",
                 pvalue_colname = "FDR",
                 display_bar_value = "Completeness",
                 number_display = 20,
                 up_down_fgsea = TRUE,
                 type = "fgsea", 
                 log_pvalue = TRUE
               ) %>% print()
               cat("\n")
               
             })

```



### **Main significant pathways - dotplots**  {.tabset}

<!-- Creates a tab with the plot for each contrast  -->

```{r, results='asis', fig.width= 8, fig.height=6}
purrr::pwalk(DEA_res,
             function(fGSEA_results_4graph, contrast, ...) {
               cat(paste0("  \n#### ", contrast , "  \n"))
               draw_dotplot_enrichment(
                 fGSEA_results_4graph,
                 xlabel = "NES",
                 pathway_colname = "ID",
                 pvalue_colname = "FDR",
                 display_dot_value = "Completeness",
                 pathway_size_colname = "pathway_size",
                 number_display = 10,
                 up_down_fgsea = TRUE,
                 type = "fgsea",
                 log_pvalue = TRUE
               ) %>% print()
               cat("\n")

             })
```

##

:::conclusion
+ Several pathways are common between the 3 treatments even though no genes were DEGs for BBB
:::

### **Pathway visualization** {.tabset} 

#### ES plot of the top 10 Up-regulated pathways {.tabset}

<!-- Creates a tab with the plots for each contrast  -->

```{r, results='asis', fig.width=12, fig.height=8 , warnings = FALSE}
purrr::pwalk(DEA_res,
             function(fGSEA_results, contrasts, ...) {
               cat("  \n  \n")
               cat(paste0("  \n##### ",
                          contrasts,
                          " {.tabset} \n"))
               if (nrow(fGSEA_results@result %>% filter(NES > 0)) > 0) {
                 purrr::map(fGSEA_results@result %>% filter(NES > 0) %>% slice_min(p.adjust, n = 10) %>% pull(Description),
                            function(desc) {
                              cat("  \n  \n")
                              cat(paste0("  \n###### ",
                                         desc,
                                         " \n"))
                              draw_running_score(fGSEA_results, pathway_name = desc) %>% print()
                              cat("\n")
                            })
                 
                 cat("\n")
                 
               }
               cat("\n")
             })
```


#### ES plot of the top 10 Down-regulated pathways {.tabset}

<!-- Creates a tab with the plots for each contrast  -->

```{r, results='asis', fig.width=12, fig.height=8 , warnings = FALSE}
purrr::pwalk(DEA_res,
             function(fGSEA_results, contrasts, ...) {
               cat("  \n  \n")
               cat(paste0("  \n##### ",
                          contrasts,
                          "\n"))
               if (nrow(fGSEA_results@result %>% filter(NES<0)) > 0) {
                 purrr::map(fGSEA_results@result %>% filter(NES < 0) %>% slice_min(p.adjust, n = 10) %>% pull(Description),
                            function(desc) {
                              cat("  \n  \n")
                              cat(paste0("  \n###### ",
                                         desc,
                                         " \n"))
                              draw_running_score(fGSEA_results, pathway_name = desc) %>% print()
                              cat("\n")
                            })
                 
                 cat("\n")
                 
               }
               cat("\n")
             })
```

#### Focus on particular pathways

```{r , fig.width = 5, fig.height = 5}

draw_running_score (DEA_res$fGSEA_results[[1]], pathway_name = 1:3)

```
