<!--------------------------------------------------------------------------------------------->
<!--                                                                                         -->
<!--                                   Enrichment analysis                                   -->
<!--                                                                                         -->
<!--------------------------------------------------------------------------------------------->

```{r, echo = FALSE}
library(msigdbr)
library(clusterProfiler)
library(fgsea)
library(enrichplot)
library(ggstance)
library(forcats)
```

# **Enrichment Analysis**

<!--------------------------------------------------------------->
<!--                                                           -->
<!--                  Pathway database selection               -->
<!--                                                           -->
<!--------------------------------------------------------------->

```{r}
# MSigDB
m_df = msigdbr(species = "Homo sapiens", category = "H")
m_list = m_df %>% split(x = .$gene_symbol, f = .$gs_name)

enrichDB <- m_df %>%
  dplyr::select(gs_name, gene_symbol)

# Reactome
# enrichDB <- msigdbr::msigdbr(species = "Homo sapiens",
#                  category = "C2",
#                  subcategory = "REACTOME") %>% select(gs_name, gene_symbol)

# KEGG
# enrichDB <- msigdbr::msigdbr(species = "Homo sapiens",
#                  category = "C2",
#                  subcategory = "KEGG") %>% select(gs_name, gene_symbol)

```

<!--------------------------------------------------------------->
<!--                                                           -->
<!--                       Fisher approach                     -->
<!--                                                           -->
<!--------------------------------------------------------------->

## **Fisher approach**

+ Fisher enrichment analysis is based on Fisher's exact test which is a hypergeometric test of over-representation applied to a contingency table comparing a list of DEGs and the list of genes of a given pathway

+ We consider, as a reference geneset, all genes available before the pre-processing, and genesets of interest based on DEGs with the FC and p-value threshold used for the differential expression analysis


```{r}
# Running Fisher test
DEA_res <- DEA_res %>% mutate(Fisher_results = purrr::map(
    DEgenes,
    ~ .x %>% pull(Genes) %>% fisher_enrichment(
    significative_genes_list = .  ,
    background_genes = rownames(eset_count),
    pathways_database = enrichDB
  )
  ))

# Calculation of the completeness of each pathway for graphical representation
completnesse_data <- completeness_calculation(
  analyzed_data = fData(eset_count),
  gene_colname_data = "GeneID",
  reference_database = enrichDB,
  pathway_colname_reference = "gs_name",
  gene_colname_reference =  "gene_symbol"
)


# adding formatted Fisher results object to DEA_res

DEA_res <- DEA_res %>% mutate(
  Fisher_results_4graph = purrr::map(
    Fisher_results,
    function(.x){
      if(is.null(.x)){
        NULL
      }else{
        .x %>% res_format_enrichment(., type = "fisher")
      }
    } 
  )
)

DEA_res <- DEA_res %>% mutate(
  Fisher_results_4graph = purrr::map(
    Fisher_results_4graph,
    function(.x){
      if(is.null(.x)){
        NULL
      }else{
        .x %>% left_join(., completnesse_data)
      }
    } 
  )
)

```


<!--------------------------------------------------------------->
<!--                                                           -->
<!--                  Pathways visualization                   -->
<!--                                                           -->
<!--------------------------------------------------------------->


## **Main significant pathways - barplots**  {.tabset}


:::info
+ The x-axis of the plot displays the geneRatio that represents the relative proportion of DEGs in a pathway

+ GeneRatio = #{genes in the overlap (DEGs geneset) $\bigcap$ (genes of the pathway)} / #{genes in the overlap (DEGs geneset) \bigcap (genes of ALL pathways of the selected database)}

+ At the end of each bar is displayed the completeness that represents the amount of genes of the pathway present in the reference gene list

+ Completeness = #{genes in the pathway} / #{genes in the reference geneset}

:::


<!-- Creates a tab with the plot for each contrast  -->

```{r, results='asis', fig.width= 8, fig.height=6}
purrr::pwalk(DEA_res ,
             function(Fisher_results_4graph, contrast, ...) {
               cat(paste0("  \n### ", contrast , "  \n"))
               if (!is.null(Fisher_results_4graph)) {
               draw_barplot_score (
                 Fisher_results_4graph,
                 xlabel = "GeneRatio",
                 pathway_colname = "ID",
                 pvalue_colname = "FDR",
                 display_bar_value = "Completeness",
                 number_display = 15,
                 up_down_fgsea = FALSE,
                 type = "fisher",
                 bar_width = 0.6, 
                 log_pvalue =  TRUE
               ) %>% print()
               }
               cat("\n")
               
             })
```



## **Main significant pathways - dotplots**  {.tabset}

<!-- Creates a tab with the plot for each contrast  -->

```{r, results='asis', fig.width= 8, fig.height=6}
purrr::pwalk(DEA_res %>% filter(contrast != "Ox40_vs_Iso"),
             function(Fisher_results_4graph, contrast, ...) {
               cat(paste0("  \n### ", contrast , "  \n"))
               if (!is.null(Fisher_results_4graph)) {
                 draw_dotplot_enrichment(
                   enrichment_results = Fisher_results_4graph,
                   xlabel = "GeneRatio",
                   pathway_size_colname = "pathway_size",
                   pathway_colname = "Description",
                   pvalue_colname = "FDR",
                   number_display = 10,
                   display_dot_value = "Completeness",
                   up_down_fgsea = TRUE,
                   type = "fisher",
                   log_pvalue = TRUE
                 )   %>% print()
               }
               cat("\n")
               
             })

```

##

:::conclusion
+ Interesting pathways can be identified for both AAA and CCC
:::

