---
title: "Code_examles"
author: "Team"
date: "03/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- Example of codes for bullet points -->

1. First item on our numbered list
    * a bullet point
    - a bullet point
    + still a bullet point
  
2. Second item on our numbered list
    * a bullet point:
        * Now with a sub-list to our sub-list
            * still with a sub-list to our sub-list
    - a bullet point
    + still a bullet point

    
    
    
    
## R Markdown


# Template to be used in rmd

## Bloc


Cette partie permet de faire des colonnes côte à côte

::::mycolumns

:::mycolumn
Here is the **first** Div.

```{r}
DT::datatable(iris)
```
:::

:::mycolumn
And this block will be put on the right:

```{r}
flextable::flextable(iris)
```
:::
::::

## Colonnes dans une boucle


```{r, results='asis'}
cat("\n")
purrr::walk(1:3, function(x){
  cat("\n")
  cat("::::mycolumns")
  cat("\n")
  cat(":::mycolumn")
  cat("\n")
  cat(knitr::knit_print(DT::datatable(iris)))
  cat("blabla")
  cat("\n")
  cat(":::")
  cat("\n")
  cat(":::mycolumn")
  cat("\n")
  cat("blabla")
  cat(knitr::knit_print(DT::datatable(iris)))
  cat("\n")
  cat(":::")
  cat("\n")
  cat("::::")
  cat("\n")

})
cat("\n")

```
